---
title: "Phase 3 — Selection-Effect Investigation"
subtitle: "Pacific Life AI Actuary Challenge 2026"
author: "Team 7"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
  pdf_document:
    toc: true
knitr:
  opts_chunk:
    echo: true
    message: false
    warning: false
    fig.align: center
    fig.width: 8
    fig.height: 5
    dpi: 150
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))

```

# Purpose

The Phase 2 modeling result was negative: adding wearable + diagnosis features to the baseline underwriter (M0) **degrades** out-of-fold log-loss and breaks calibration. To make the recommendation defensible we need to explain *why* the apparent crude wearable benefit (8.34% -> 5.82% mortality) doesn't survive inclusion in a model that already has age, BMI, smoker, and diagnosed conditions.

We test five things in this notebook:

1. **Propensity model** — *who* chooses to wear a device?
2. **Covariate balance** before and after inverse-propensity-treatment-weighting.
3. **IPTW marginal effect** of wearable on mortality.
4. **Engagement dose-response** — raw and age-stratified.
5. **Negative-control falsification** — variables that *cannot* causally
   affect mortality (operating system, dominant hand) but show how much
   confounding remains.

# Setup

```{r packages}
suppressPackageStartupMessages({
  library(dplyr); library(readr); library(ggplot2); library(tidyr)
  library(WeightIt); library(cobalt); library(scales); library(patchwork)
  library(broom)
})

set.seed(42)
dir.create("output/figures", showWarnings = FALSE, recursive = TRUE)
dir.create("output/tables",  showWarnings = FALSE, recursive = TRUE)
```

```{r load-data}
df <- read_csv("data/processed/features_train.csv", show_col_types = FALSE)
glimpse(select(df, ID, wearable, engagement_level, age, bmi, outcome))
```

# 1. Propensity Model — who chooses a wearable?

If wearable adopters are systematically different from non-adopters, any
mortality difference between the groups is partly *selection* and partly
(possibly zero) *causal*. The propensity model estimates the probability of
adopting based only on baseline traits.

```{r propensity-fit}
prop_vars = c("age", "bmi", "smoker", "weekly_exercise",
               "alcohol_consumption_frequency",
               "family_medical_diabetes", "family_medical_cancer",
               "family_medical_dementia", "family_medical_heart_disease",
               "diagnosed_diabetes", "diagnosed_lung_cancer",
               "diagnosed_other_cancer", "diagnosed_sleep_apnea",
               "diagnosed_afib", "diagnosed_hypertension",
               "diagnosed_dementia")

prop_formula = as.formula(
  paste("wearable ~ sex + age_bin +", paste(prop_vars, collapse = " + "))
)

prop_fit = glm(prop_formula, family = binomial(), data = df)
df$pscore = predict(prop_fit, type = "response")
```

```{r propensity-coefs}
prop_tbl = tidy(prop_fit, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  arrange(p.value)

write_csv(prop_tbl, "output/tables/phase3_propensity_coefs.csv")

knitr::kable(head(prop_tbl, 10), digits = 3,
             caption = "Top 10 wearable-adoption predictors (odds ratios)")
```

**Interpretation.** `weekly_exercise` has an odds ratio of ~1.69 — each
additional day per week of exercise nearly doubles the odds of adopting a
wearable. `alcohol_consumption_frequency` and most diagnosed-condition flags
push the other way. **Adopters are systematically healthier on observable
lifestyle dimensions.**

```{r overlap-plot, fig.cap="Propensity-score distribution by wearable status"}
p_overlap = ggplot(df, aes(x = pscore, fill = factor(wearable))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "#d7191c", "1" = "#2c7bb6"),
                    labels = c("No wearable", "Wearable"), name = NULL) +
  labs(title = "Propensity-score overlap",
       subtitle = "Wearable adopters are systematically different",
       x = "P(wearable = 1 | baseline traits)", y = "Density")
  theme_minimal(base_size = 12)

ggsave("output/figures/phase3_pscore_overlap.png",
       p_overlap, width = 7, height = 4.5, dpi = 200)
p_overlap

```

# 2. IPTW — covariate balance before vs. after weighting

`WeightIt::weightit` builds stabilized inverse-propensity weights so the
treated and control groups look comparable on observable covariates. We
visualize this with a `cobalt` love plot.

```{r iptw-fit}
W = weightit(prop_formula, data = df, method = "glm",
              estimand = "ATE", stabilize = TRUE)
df$ipw = W$weights
summary(W$weights)
```

```{r love-plot, fig.cap="Standardized mean differences before vs. after IPTW"}
love.plot(W, binary = "std", thresholds = c(m = 0.1),
          title = "Covariate balance: unweighted vs. IPTW",
          colors = c("#e84118", "#1e3799"),
          shapes = c(16, 17))

png("output/figures/phase3_balance_loveplot.png",
    width = 1400, height = 900, res = 180)
print(love.plot(W, binary = "std", thresholds = c(m = 0.1),
                title = "Covariate balance: unweighted vs. IPTW",
                colors = c("#e84118", "#1e3799"),
                shapes = c(16, 17)))
invisible(dev.off())
```

# 3. Marginal effect — crude vs. IPTW

```{r marginal-rates}
crude_rates <- df |>
  group_by(wearable) |>
  summarise(n = n(), deaths = sum(outcome), rate = mean(outcome))

iptw_rates <- df |>
  group_by(wearable) |>
  summarise(n = n(),
            rate_ipw = sum(outcome * ipw) / sum(ipw))

est_tbl <- crude_rates |>
  left_join(iptw_rates |> select(wearable, rate_ipw), by = "wearable")

write_csv(est_tbl, "output/tables/phase3_marginal_estimates.csv")
knitr::kable(est_tbl, digits = 4,
             caption = "Mortality (crude vs. IPTW-adjusted)")
```

```{r ate-summary}
crude_diff = diff(crude_rates$rate)
iptw_diff = diff(iptw_rates$rate_ipw)
crude_rr = crude_rates$rate[2] / crude_rates$rate[1]
iptw_rr = iptw_rates$rate_ipw[2] / iptw_rates$rate_ipw[1]

ate_tbl = tibble(
  estimand = c("abs risk diff (wear - no_wear)",
               "relative risk (wear / no_wear)"),
  crude = c(crude_diff, crude_rr),
  iptw  = c(iptw_diff, iptw_rr)
)

write_csv(ate_tbl, "output/tables/phase3_ate_summary.csv")
knitr::kable(ate_tbl, digits = 4,
             caption = "Wearable effect estimates: crude vs. IPTW")
```

**Important nuance.** The IPTW estimate is essentially identical to the
crude estimate. That tells us *observable* covariates do **not** explain the
wearable–mortality association. Two interpretations remain:

- **(a) Real protective effect** — possible but uncomfortably large for a
  wellness intervention.
- **(b) Unobserved confounding** — adopters likely differ on wealth, SES,
  healthcare access, and behavioral traits we cannot measure.

The negative-control test in Section 5 lets us distinguish these.

# 4. Engagement dose-response

```{r dose-response, fig.width=11, fig.height=4.5, fig.cap="Mortality by engagement (raw and age-stratified)"}
dose_raw = df |>
  group_by(engagement_level) |>
  summarise(n = n(), rate = mean(outcome),
            se = sqrt(rate * (1 - rate) / n))

p_dose_raw = ggplot(dose_raw,
                     aes(x = factor(engagement_level), y = rate)) +
  geom_col(fill = "#1e3799", alpha = 0.85) +
  geom_errorbar(aes(ymin = pmax(0, rate - 1.96 * se),
                    ymax = rate + 1.96 * se),
                width = 0.15, color = "gray30") +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)),
            vjust = -1.6, size = 3.2) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Mortality by engagement tier (raw)",
       subtitle = "Tier 5 reverses the trend — selection signal",
       x = "Engagement level (0 = no wearable)", y = "10-yr mortality rate") +
  theme_minimal(base_size = 12)

dose_age <- df |>
  group_by(age_bin, engagement_level) |>
  summarise(n = n(), rate = mean(outcome), .groups = "drop") |>
  filter(n >= 20)

p_dose_age = ggplot(dose_age,
                     aes(x = factor(engagement_level), y = rate,
                         group = age_bin, color = age_bin)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_brewer(palette = "Set1", name = "Age bin") +
  labs(title = "Engagement-mortality, stratified by age",
       subtitle = "Age explains most of apparent benefit",
       x = "Engagement level", y = "Mortality rate") +
  theme_minimal(base_size = 12)

combined = p_dose_raw + p_dose_age + plot_layout(widths = c(1, 1.2))
ggsave("output/figures/phase3_dose_response.png",
       combined, width = 11.5, height = 4.5, dpi = 200)
combined
```

# 5. Negative-control falsification

If the wearable–mortality link is causal, then variables with no plausible causal path to mortality (operating system, dominant hand) should show **flat** mortality across their levels. If they don't, residual confounding remains.

```{r neg-controls}
neg_vars = c("dominant_hand", "os")
neg_results <- list()

for (v in neg_vars) {
  if (v %in% names(df)) {
    sub = df |>
      filter(!is.na(.data[[v]])) |>
      group_by(across(all_of(v))) |>
      summarise(n = n(), rate = mean(outcome), .groups = "drop") |>
      filter(n >= 50)
    sub$variable = v
    names(sub)[1] = "level"
    neg_results[[v]] = sub
  }
}

neg_df = bind_rows(neg_results)
write_csv(neg_df, "output/tables/phase3_negative_controls.csv")

knitr::kable(neg_df, digits = 4,
             caption = "Mortality by 'placebo' variables (should be flat)")
```

**Smoking gun.** Mortality varies by **operating system version** — iOS 18 users die at 4.88%, Android 13 users at 7.87%. OS cannot causally affect mortality, so this 3-percentage-point gap is **pure confounding** (plausibly socioeconomic: newest/most-expensive devices correlate with wealth, healthcare access, and survival). Any wearable feature is going to inherit this confounding. That is why the M1 model in Phase 2 fails to generalize.

# 6. Persist propensity scores & weights for downstream use

```{r persist}
df |>
  select(ID, wearable, engagement_level, pscore, ipw, outcome) |>
  write_csv("output/tables/phase3_pscore_weights.csv")

cat("Phase 3 complete. Outputs:\n",
    "  - output/figures/phase3_pscore_overlap.png\n",
    "  - output/figures/phase3_balance_loveplot.png\n",
    "  - output/figures/phase3_dose_response.png\n",
    "  - output/tables/phase3_propensity_coefs.csv\n",
    "  - output/tables/phase3_marginal_estimates.csv\n",
    "  - output/tables/phase3_ate_summary.csv\n",
    "  - output/tables/phase3_negative_controls.csv\n",
    "  - output/tables/phase3_pscore_weights.csv\n", sep = "")
```

# Take-aways for the deck

| Finding | Where it lives in the deck |
|---|---|
| Adopters self-select on `weekly_exercise` (OR ≈ 1.69) | Slide 6 (selection effects) |
| IPTW barely changes the wearable effect | Slide 6 (residual unobserved confounding) |
| Engagement tier 5 reverses the dose-response | Slide 3 (cohort design) |
| OS version predicts mortality (placebo violation) | Slide 6 (smoking gun) |
| Wearable signal cannot be cleanly identified | Slide 8 (recommendation) |
